Algorithms for Cost- and Deadline-Constrained Provisioning for Scientific Workflow Ensembles in IaaS Clouds

### 2013-08-30 paper/paper.tex
Update to paper after Ewa's comments

% can use linebreaks \\ within to get better formatting as desired
\title{Cost- and Deadline-Constrained Provisioning for Scientific Workflow
Ensembles in IaaS Clouds}

@@

\address[agh]{AGH University of Science and Technology, Dept. of Computer Science, Krakow, Poland} 
\address[isi]{USC Information Sciences Institute, 4676 Admiralty Way, Marina del Rey, CA, USA} 
\address[nd]{University of Notre Dame, Center for Research Computing, 111 ITC, Notre Dame, IN, USA}

@@

Large-scale applications expressed as scientific workflows are
often grouped into ensembles of inter-related workflows. In this paper, we
address a new and important problem concerning the efficient management of
such ensembles under budget and deadline constraints on Infrastructure-
as-a-Service (IaaS) clouds. We discuss, develop, and assess algorithms based
on static and dynamic strategies for both task scheduling and resource
provisioning. We perform the evaluation via simulation using a set of
scientific workflow ensembles with a broad range of budget and deadline
parameters, taking into account uncertainties in task runtime estimations,
provisioning delays, and failures. We find that the key factor determining the
performance of an algorithm is its ability to decide which workflows in an
ensemble to admit or reject for execution. Our results show that an admission
procedure based on workflow structure and estimates of task runtimes can
significantly improve the quality of solutions.

@@

\begin{keyword}
scientific workflows, workflow ensembles, cloud computing, resource provisioning
\end{keyword}

@@

Scientific workflows, usually represented as Directed Acyclic Graphs (DAGs),
are an important class of applications that lead to challenging problems in
resource management on grid and utility computing systems. Workflows for large
computational problems are often composed of several inter-related workflows
grouped into {\em ensembles}.  Workflows in an ensemble typically have a
similar structure, but they differ in their input data, number of tasks, and
individual task sizes.

There are many applications that require scientific workflow ensembles.
CyberShake~\cite{Callaghan2011}, for example, uses ensembles to generate
seismic hazard maps.  Each workflow in a CyberShake ensemble generates a
hazard curve for a particular geographic location, and several hazard curves
are combined to create a hazard map.  In 2009 CyberShake was used to generate
a map that required an ensemble of 239 workflows.  Similarly, users of
Montage~\cite{Deelman2008} often need several workflows  with different
parameters to generate a set of image mosaics that can be combined  into a
single, large mosaic. The Galactic Plane ensemble, which generates several
mosaics of the  entire sky in different wavelengths, consists of 17 workflows,
each of which contains 900 sub-workflows.  Another ensemble example is the
Periodograms application~\cite{Vockler2011}, which searches for extrasolar
planets by detecting periodic dips in the light intensity of their host star.
Due to the large scale of the input data, this application is often split up
into multiple batches processed by different workflows. Additional workflows
are created to run the analysis using different parameters. A recent analysis
of Kepler satellite data required three ensembles of 15 workflows.

@@

In this paper, we aim to gain insight into resource management challenges when
executing scientific workflow ensembles on clouds. We address a
new and important problem of maximizing the number of completed workflows from
an ensemble under both budget and deadline constraints. The motivation for this
work is to answer the fundamental question of concern to a researcher: How much
computation can be completed given the limited budget and timeframe of a
research project?

The main contributions of this paper are:
\begin{itemize}
  \item we define the problem of scheduling prioritized workflow ensembles
  under budget and deadline constraints,
  \item we analyze and develop several dynamic (online) and static (offline) 
  algorithms for task scheduling and resource provisioning that rely on workflow 
  structure information (critical paths and workflow levels) and estimates of 
  task runtimes,
  \item we evaluate these algorithms using a simulator based on 
  CloudSim~\cite{Calheiros2011}, which models the infrastructure and the 
  application,taking into account uncertainties in task runtime estimates, 
  provisioning delays, and failures, 
  \item we discuss the performance of the algorithms on a set of
  synthetic workflow ensembles based on important, real scientific
  applications, using a broad range of different application scenarios and
  varying constraint values.
\end{itemize}

This paper is an extension of our earlier conference
publication~\cite{Malawski-SC12}.
Here we present a more detailed performance evaluation, including the discussion
of results based on deadline and budget constraints. We also report on new
results and discussion of task granularity in the context of the resource
billing cycle, which is important aspect of scheduling and resource provisioning
on cloud infrastructures.

@@

\section{Related Work}
\label{sec:related}
General policy and rule-based approaches to dynamic provisioning (e.g. Amazon
Auto Scaling~\cite{Autoscaling} and
RightScale~\cite{RightScale}) allow the size
of a resource pool to be adjusted based on infrastructure and
application metrics. A typical infrastruc\-ture-specific metric is system load,
whereas application-speci\-fic metrics include response time and length of a task or
of a request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.
Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniques. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of workflow-aware heuristics that cannot be applied to
independent tasks.

Our work is related to the strategies for deadline-con\-strained
cost-minimization workflow scheduling, developed for utility grid systems. However, our problem is
different from \cite{Yu2005} and \cite{Abrishami2013158} in that we consider ensembles of workflows in
IaaS clouds, which allow one to provision resources on a per-hour billing model,
rather than utility grids, which allow one to choose from a pool of existing
resources with a per-job billing model. Our work is also different from
cloud-targeted autoscaling solution~\cite{Mao2011} in that we consider ensembles
of workflows rather than unpredictable workloads containing workflows. We also consider budget constraints
rather than cost minimization as a goal. In other words, we assume that there is
more work to be done than the available budget, so some work must be rejected.
Therefore, cost is not something we optimize (i.e. an objective), but rather a constraint.

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan2010,Dongarra2007}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints. Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in the scenarios we consider in this paper. Our work can also
be regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

\underline{Resource Model}  We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2),
where virtual machine (VM) instances may be provisioned on-demand and are billed by
the hour, with partial hours being rounded up. Although there may be
heterogeneous VM types with different amounts of CPU, memory, disk space, and
I/O, for this paper we focus on a single VM type because we assume that for most
applications there will typically be only one or two VM types with the best
price/performance ratio for the application~\cite{Juve2009}. We assume that
a submitted task has exclusive access to a VM instance and that there is no
preemption. We also assume that there is a delay between the time that a new VM
instance is requested and when it becomes available to execute tasks.

\underline{Application Model} The target applications are ensembles of
scientific workflows that can be modeled as DAGs, where the nodes in the graph
represent computational tasks, and the edges represent data- or control-flow
dependencies between the tasks. We assume that runtime estimates for the
workflow tasks are known, but that they are not perfect and may vary based on
a uniform distribution of $\pm p\%$.

@@

\underline{Performance Metric}
\label{sec:perf_metric}
In order to precisely define the objective of the algrithms it is necessary to
introduce a metric that can be used to score the performance of the different
algorithms on a given problem (ensemble, budget, and deadline). The simplest
approach is to count the number of workflows in the ensemble that each algorithm
is able to complete within the budget before the deadline, but this metric does
not account for the priority-based utility function specified by the user. Using
the counting approach, a less efficient algorithm may be able to complete a
large number of low-priority workflows by executing the smallest workflows
first. In order to account for the priority, we use an exponential
scoring defined as:
$$ Score(e) = \sum_{w~\in~Completed(e)}{2^{-Priority(w)}} $$
where $Completed(e)$ is the set of workflows in ensemble $e$ that was completed
by the algorithm, and $Priority(w)$ is the priority of workflow $w$ such that
the highest-priority workflow has $Priority(w)=0$, the next highest workflow has
$Priority(w)=1$, and so on. This exponential scoring function gives the highest
priority workflow a score that is higher than all the lower- priority workflows
combined:
$$ 2^{-p} > \sum_{i~=~p+1,~\ldots}2^{-i} $$
This scoring is consistent with our definition of the problem, which is to
complete as many workflows as possible, according to their priorities, given a
set budget and deadline.

@@

DPDS guarantees that tasks from lower priority workflows are always deferred
when higher-priority tasks are available, but lower-priority tasks can still
occupy idle VMs when higher-priority tasks are not available. Since there is
no preemption, long-running low-priority tasks may delay the execution of
higher-priority tasks. In addition, tasks from low priority workflows may be
executed even though there is no chance that those workflows will be completed
within the current budget and deadline. Fig.~\ref{fig:algorithm-example-dpds}
shows an example schedule generated using the DPDS algorithm. The figure
illustrates how tasks from lower priority workflows backfill idle VMs when
tasks from higher priority workflows are not available.

@@

The previous dynamic (online) algorithms make provisioning and scheduling
decisions at runtime. By contrast, the SPSS algorithm creates a provisioning
and scheduling plan before running any workflow tasks. This enables SPSS to
start only those workflows that it knows can be completed given the deadline and
budget constraints, and eliminates any waste that may be allowed by the dynamic
algorithms. However, the disadvantage of SPSS is that it is sensitive to dynamic
changes in the environment (see Sections~\ref{sec:delays} and
\ref{sec:variances}).

The approach used by SPSS is to plan each workflow in the ensemble in priority
order, rejecting any workflows that would exceed the dedline or budget. Once the
plan is complete, the VMs are provisioned and tasks are executed according to
the schedules given by the plan.

The disadvantage of the static planning approach used by SPSS is that it is
sensitive to dynamic changes in the environment and the application that may
disrupt its carefully constructed plan. For example, if there are provisioning
delays, or if the runtime estimates for the tasks are inaccurate, then workflow
execution may diverge from the plan. This issue will be discussed further in
Sections~\ref{sec:delays} and \ref{sec:variances}.

@@

An example schedule generated by SPSS (Fig.~\ref{fig:algorithm-example-spss})
shows how SPSS tends to start many workflows in parallel, running each workflow
over a longer period of time on only a few VMs to minimize cost. In comparison,
the dynamic algorithms tend to run one workflow at a time across many VMs in
parallel.

@@

To evaluate and compare the three proposed algorithms, we developed a cloud
workflow simulator~\cite{CloudWorkflowSimulator} based on CloudSim~\cite{Calheiros2011}. Our
simulation model consists of Cloud, VM and WorkflowEngine entities. The Cloud entity starts and
terminates VM entities using an Amazon EC2-like API. VM entities simulate the
execution of individual tasks, including randomized variations in runtime. The
WorkflowEngine entity manages the scheduling of tasks and the provisioning of 
VMs based on the chosen algorithm. We assume that the VMs have a single core
and execute tasks sequentially. 
Although CloudSim provides a more advanced infrastructure model which includes
time-sharing and space-sharing policies, we do not use these features since we
are interested mainly in the execution of tasks on VMs and high-level workflow
scheduling and provisioning. 
The simulator reads workflow description files in
a modified version of the DAX format used by the Pegasus Workflow Management
System~\cite{Deelman2005}. 
%The modified file format was created for the
%synthetic workflow generator developed by Bharathi, et al~\cite{Bharathi2008}.

@@

Pareto ensembles contain a small number of larger workflows and a large number
of smaller workflows. Their sizes  
%in a Pareto ensemble 
are chosen according to a Pareto distribution. The distribution was modified so
that the number of large workflows (of size $\geq$ 900) is increased by a small amount to
produce a ``heavy-tail''. This causes Pareto ensembles to have a slightly larger
number of large workflows, which reflects behavior commonly observed in many
computational workloads. 
An example of the distribution of workflow sizes that
occurs in a Pareto ensemble is shown in Fig.~\ref{fig:ensemble-pareto}.

\begin{figure}[ht] 
    \centering
    \includegraphics[width=0.6\columnwidth]{figures/ensemble-pareto}
    \caption[Histogram of workflow sizes in Pareto ensembles]{Histogram of
    workflow sizes in Pareto ensembles. Workflow size is measured in number of
    tasks.}
    \label{fig:ensemble-pareto}
\end{figure}

@@

@@

Fig.~\ref{fig:distributions} shows the percentage of simulations for which
each algorithm achieved the highest score for a given ensemble type. This
experiment was conducted using all five applications, with all five types of
ensembles. For each application and ensemble type, 10 random ensembles of 50
workflows each were created. Each ensemble was simulated with all three
algorithms using 10 budgets and 10 deadlines (1,000 simulations per
application, ensemble type, and algorithm, or 75,000 in total). The best scores percentage is
computed by counting the number of times that a given algorithm achieved the
highest score and dividing by 1000. Note that it is possible for multiple
algorithms to get the same high score (to tie), so the numbers do not
necessarily add up to 100\%. The sum is much higher than 100\% in cases where
the dynamic algorithms perform relatively well because DPDS and WA-DPDS, which
are very similar algorithms, often get the same high score.

\begin{figure*}[ht]
    \centering
    \includegraphics[width=\textwidth]{run-finish-variations-test-0-output-distributions}
    \caption{Percentage of high scores achieved by each algorithm on different ensemble types for all five applications. C = Constant ensembles, PS = Pareto Sorted ensembles, PU = Pareto Unsorted ensembles, US = Uniform Sorted ensembles, and UU = Uniform Unsorted ensembles.} 
    \label{fig:distributions}
\end{figure*}

There are several interesting things to notice about
Fig.~\ref{fig:distributions}. The first is that, in most cases, SPSS
significantly outperforms both dynamic algorithms (DPDS and WA-DPDS). This is
attributed to the fact that SPSS is able to make more intelligent scheduling
and provisioning decisions because it has the opportunity to compare different
options and choose the one that results in the best outcome. In comparison,
the dynamic algorithms are online algorithms and are not able to project into
the future to weigh the outcomes of their choices.

@@

\subsection{Results vs. deadline and budget}

In order to give more details about the performance of the algorithms, in this
section we show how the results depend on the deadline and budget constraints.
As the performance metric we use the number of completed workflows of the
ensemble, since the exponential score is not suitable for showing trends.
We use all deadline and budget parameters as defined in
Section~\ref{sec:exp-parameters} and normalize them to the range from 0 to 1 to
facilitate plotting.
For each normalized deadline, we compute the mean number of completed workflows,
averaged over all budgets and random seeds, so each point on the plot is based
on 100 different simulation runs. In Fig.~\ref{fig:completed-deadlines}
and~\ref{fig:completed-budgets} we present the results of CyberShake and LIGO
applications, as representatives of fine-grained and coarse-grained workflows.
Similar results were obtained for other applications.

Fig.~\ref{fig:completed-deadlines} shows the results depending on normalized
deadline. It can be observed, that with increasing deadline, the number of
completed worklfows grows steeply and then stabilizes. This means that for
longer deadlines it is usually not possible to complete more workflows due to
the budget constraint. For short deadlines, the algorithms need to allocate more
VM instances in parallel, making it more difficult to utilize the resources
efficiently, which results in lower numbers of completed workflows.

We can also observe that the algorithms are able to complete more workflows in
the ensembles that have priorities assigned randomly (unsorted), in comparison
to the ensembes where the priorities are assigned based on workflow size
(sorted). This results from the fact that in the unsoted distributions the large
workflows often have low priorities, so they can be rejected by the algoritms,
thus making more space in the schedule for smaller high-priority workflows.

One interesting observation from Fig.~\ref{fig:completed-deadlines} is that for
CyberShake application the result curves have a sawtooth shape. This results
from the combination of two effects. The first comes from the fact that dynamic
algorithms perform better when the deadline is close to the full hour, due to
the provisioning mechanism that tries to shut down unused VM instances before a
full hour. The second one is that when the deadline increases, it may become
possible to admit a larger workflow with higher priority instead of several
smaller ones with lower priority. That is why the number of completed workflows
may decrese, but the exponential score (utility function) based on priorities
will be higher. These effects are visible for CyberShake application, for which
the maximum deadline is 3 hours, but they become less important for LIGO
application, for which the maximum deadline is 20 hours. See also the discussion
on task granularity in Setion~\ref{sec:granularity}.

@@

\begin{figure*}[t!]
    \centering
    \includegraphics[width=0.9\textwidth]{run-finish-variations-test-0-output-deadlines-app-dist-selected}    
    \caption{Number of completed workflows from the ensemble depending on normalized deadline.}
    \label{fig:completed-deadlines}
\end{figure*}

\begin{figure*}[t!]
    \centering
    \includegraphics[width=0.9\textwidth]{run-finish-variations-test-0-output-budgets-app-dist-selected}    
    \caption{Number of completed workflows from the ensemble depending on normalized budget.}
    \label{fig:completed-budgets}
\end{figure*}

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.9\columnwidth]{run-stretching-test-output-scaling-CYBERSHAKE}
    \includegraphics[width=0.9\columnwidth]{run-stretching-test-output-scaling-MONTAGE}

    \caption{Percentage of high scores achieved by each algorithm for CyberShake
    and Montage ensembles when task runtime is scaled by a constant factor. A
    scaling factor of $x$ means that the runtime of tasks in each workflow was
    multiplied by $x$.}
    \label{fig:stretching}
\end{figure}

@@

Fig.~\ref{fig:stretching} shows the relative performance of the algorithms as
the scaling factor is increased from 1 to 16 for CyberShake  and Montage
applications. Each data point in the figure represents 500 simulations (5
ensembles with 10 budgets and 10 deadlines). The best scores percent was
calculated the same way as it was for Fig.~\ref{fig:distributions}.

>>>>

% can use linebreaks \\ within to get better formatting as desired
\title{Algorithms for Cost- and Deadline-Constrained Provisioning for Scientific Workflow
Ensembles in IaaS Clouds}

@@

\address[agh]{AGH University of Science and Technology, Dept. of Computer Science, Al. Mickiewicza 30, Krakow, Poland} 
\address[isi]{USC Information Sciences Institute, 4676 Admiralty Way, Marina del Rey, CA, USA} 
\address[nd]{University of Notre Dame, Center for Research Computing, 111 ITC, Notre Dame, IN, USA}

@@

Large-scale applications expressed as scientific workflows are often grouped
into ensembles of inter-related workflows. In this paper, we address a new and
important problem concerning the efficient management of such ensembles under
budget and deadline constraints on Infrastructure- as-a-Service (IaaS) clouds.
IaaS clouds are characterized by on-demand resource provisioning capabilities
and a pay-per-use model. We discuss, develop, and assess algorithms based on
static and dynamic strategies for both task scheduling and resource
provisioning. We perform the evaluation via simulation using a set of scientific
workflow ensembles with a broad range of budget and deadline parameters, taking
into account task granularity, uncertainties in task runtime estimations,
provisioning delays, and failures. We find that the key factor determining the
performance of an algorithm is its ability to decide which workflows in an
ensemble to admit or reject for execution. Our results show that an admission
procedure based on workflow structure and estimates of task runtimes can
significantly improve the quality of solutions.

@@

\begin{keyword}
scientific workflows; workflow ensembles; cloud computing; resource provisioning
\end{keyword}

@@

Large-scale scientific workflows, usually represented as Directed Acyclic Graphs
(DAGs), are an important class of applications that lead to challenging problems
in resource management on grid and cloud infrastructures. What becomes more
important, workflows for large computational problems are often composed of
several inter-related workflows grouped into {\em ensembles}.  Workflows in an
ensemble typically have a similar structure, but they differ in their input
data, number of tasks, and individual task sizes.

There are many applications that require scientific workflow ensembles.
CyberShake~\cite{Callaghan2011}, for example, uses ensembles to generate seismic
hazard maps.  Each workflow in a CyberShake ensemble generates a hazard curve
for a particular geographic location, and several hazard curves are combined to
create a hazard map.  In the recent study of 2013 CyberShake generates a map that
requires an ensemble of 286 workflows.  Similarly, users of
Montage~\cite{Deelman2008} often need several workflows  with different
parameters to generate a set of image mosaics that can be combined  into a
single, large mosaic. The Galactic Plane ensemble, which generates several
mosaics of the  entire sky in different wavelengths, consists of 17 workflows,
each of which contains 900 sub-workflows.  Another ensemble example is the
Periodograms application~\cite{Vockler2011}, which searches for extrasolar
planets by detecting periodic dips in the light intensity of their host star.
Due to the large scale of the input data, this application is often split up
into multiple batches processed by different workflows. Additional workflows are
created to run the analysis using different parameters. A recent analysis of
Kepler satellite data required three ensembles of 15 workflows.

@@

In this paper, we aim to gain insight into resource management challenges when
executing scientific workflow ensembles on clouds. We address a
new and important problem of maximizing the number of completed workflows from
an ensemble under both budget and deadline constraints. The motivation for this
work is to answer the fundamental question of concern to a researcher: {\em How much
computation can be completed given the limited budget and timeframe of a
research project?}

The main contributions of this paper are:
\begin{itemize}
  \item we define the problem of scheduling prioritized workflow ensembles
  under budget and deadline constraints,
  \item we analyze and develop several dynamic (online) and static (offline) 
  algorithms for task scheduling and resource provisioning that rely on workflow 
  structure information (critical paths and workflow levels) and estimates of 
  task runtimes,
  \item we evaluate these algorithms using a simulator based on 
  CloudSim~\cite{Calheiros2011}, which models the infrastructure and the 
  application, taking into account uncertainties in task runtime estimates, 
  provisioning delays, and failures, 
  \item we discuss the performance of the algorithms on a set of
  synthetic workflow ensembles based on important, real scientific
  applications, using a broad range of different application scenarios and
  varying constraint values.
\end{itemize}

This paper is an extension of our earlier conference
publication~\cite{Malawski-SC12}.
Here we present a more detailed performance evaluation, including the discussion
of results based on varying deadline and budget constraints. We also report on new
results and discussion of task granularity in the context of the resource
billing cycle, which is an important aspect of scheduling and resource provisioning
on cloud infrastructures.

@@

\section{Related Work}
\label{sec:related}
General policy-based and rule-based approaches to dynamic provisioning (such as Amazon
Auto Scaling~\cite{Autoscaling} and
RightScale~\cite{RightScale}) allow the size
of a resource pool to be adjusted based on infrastructure and
application metrics. A typical infrastruc\-ture-specific metric is system load,
whereas application-speci\-fic metrics include response time and length of a task or
of a request queue. It is possible to set thresholds and limits to tune the behavior
of these autoscaling systems, but no support for complex applications is provided.
Policy-based approaches for scientific workloads (e.g. \cite{Marshall2010,
Kim2011}) also allow to scale the cloud resource pool or to extend the
capabilities of clusters using cloud-burst techniques. Our approach is different
in that we consider workflows, while policy based approaches typically consider
bags of independent tasks or unpredictable batch workloads. This enables us to
take advantage of workflow-aware heuristics that cannot be applied to
independent tasks.

Our work is related to the strategies for deadline-con\-strained
cost-minimization workflow scheduling, developed for utility grid and cloud
systems. However, our problem is different from
\cite{Yu2005},\cite{Abrishami2013158} and~\cite{Young2013} in that we consider
ensembles of workflows instead of individual workflows.
Our work is also different from cloud-targeted autoscaling
solution~\cite{Mao2011} in that we consider ensembles of workflows rather than
unpredictable workloads containing workflows. We also consider budget
constraints rather than cost minimization as a goal. In other words, we assume
that there is more work to be done than the available budget, so some work must
be rejected.
Therefore, cost is not something we optimize (i.e. an objective), but rather a
constraint.

This work is related to bi-criteria scheduling and multi-criteria scheduling of
workflows~\cite{Wieczorek2009,Prodan2010,Dongarra2007,Durillo2013}. These approaches are
similar to ours in that we have two scheduling criteria: cost and makespan. The
challenge in multi-criteria scheduling is to derive an objective function that
takes into account all of the criteria. In our case one objective (amount
of work completed) is subject to optimization, whereas time and cost are
treated as constraints.

Other approaches~\cite{Talukder2009,Pandey2010} use
metaheuristics that usually run for a long time before producing good results,
which makes them less useful in the scenarios we consider in this paper. Our work can also
be regarded as an extension of the budget-constrained workflow scheduling
\cite{Sakellariou2007} in the sense that we are dealing with workflow ensembles
and the deadline constraint is added.

@@

\subsection{Resource Model}  
We assume a resource model similar to Amazon's Elastic Compute Cloud (EC2),
where virtual machine (VM) instances may be provisioned on-demand and are billed by
the hour, with partial hours being rounded up. Although there may be
heterogeneous VM types with different amounts of CPU, memory, disk space, and
I/O, for this paper we focus on a single VM type because we assume that for most
applications there will typically be only one or two VM types with the best
price/performance ratio for the application~\cite{Juve2009}. We assume that
a submitted task has exclusive access to a VM instance and that there is no
preemption. We also assume that there is a delay between the time that a new VM
instance is requested and when it becomes available to execute tasks.

\subsection{Application Model} 
The target applications are ensembles of
scientific workflows that can be modeled as DAGs, where the nodes in the graph
represent computational tasks, and the edges represent data- or control-flow
dependencies between the tasks. We assume that runtime estimates for the
workflow tasks are known, but that they are not perfect and may vary based on
a uniform distribution of $\pm p\%$.

@@

\subsection{Performance Metric}
\label{sec:perf_metric}
In order to precisely define the objective of the algrithms it is necessary to
introduce a metric that can be used to score the performance of the different
algorithms on a given problem (ensemble, budget, and deadline). The simplest
approach is to count the number of workflows in the ensemble that each algorithm
is able to complete within the budget before the deadline, but this metric does
not account for the priority-based utility function specified by the user. Using
the counting approach, a less efficient algorithm may be able to complete a
large number of low-priority workflows by executing the smallest workflows
first. In order to account for the priority, we use an exponential
scoring defined as:
$$ Score(e) = \sum_{w~\in~Completed(e)}{2^{-Priority(w)}} $$
where $Completed(e)$ is the set of workflows in ensemble $e$ that was completed
by the algorithm, and $Priority(w)$ is the priority of workflow $w$ such that
the highest-priority workflow has $Priority(w)=0$, the next highest workflow has
$Priority(w)=1$, and so on. This exponential scoring function gives the highest
priority workflow a score that is higher than all the lower- priority workflows
combined:
$$ 2^{-p} > \sum_{i~=~p+1,~\ldots}2^{-i} $$
This scoring is consistent with our definition of the problem, which is to
complete as many workflows as possible, according to their priorities, given a
set budget and deadline.

@@

DPDS guarantees that tasks from lower priority workflows are always deferred
when higher-priority tasks are available, but lower-priority tasks can still
occupy idle VMs when higher-priority tasks are not available. Since there is
no preemption, long-running low-priority tasks may delay the execution of
higher-priority tasks. In addition, tasks from low priority workflows may be
executed even though there is no chance that those workflows will be completed
within the current budget and deadline. Figure~\ref{fig:algorithm-example-dpds}
shows an example schedule generated using the DPDS algorithm. The figure
illustrates how tasks from lower priority workflows backfill idle VMs when
tasks from higher priority workflows are not ready for execution.

@@

The previous dynamic (online) algorithms make provisioning and scheduling
decisions at runtime. By contrast, the SPSS algorithm creates a provisioning
and scheduling plan before running any workflow tasks. This enables SPSS to
start only those workflows that it knows can be completed given the deadline and
budget constraints, and eliminates any waste that may be allowed by the dynamic
algorithms. However, the disadvantage of SPSS is that it is sensitive to dynamic
changes in the environment (see Sections~\ref{sec:variances} and
\ref{sec:delays}).

The approach used by SPSS is to plan each workflow in the ensemble in priority
order, rejecting any workflows that would exceed the deadline or budget. Once the
plan is complete, the VMs are provisioned and tasks are executed according to
the schedules given by the plan.

The disadvantage of the static planning approach used by SPSS is that it is
sensitive to dynamic changes in the environment and the application that may
disrupt its carefully constructed plan. For example, if there are provisioning
delays, or if the runtime estimates for the tasks are inaccurate, then workflow
execution may diverge from the plan. This issue will be discussed further in
Sections~\ref{sec:variances} and \ref{sec:delays}.

@@

An example schedule generated by SPSS (Figure~\ref{fig:algorithm-example-spss})
shows how SPSS tends to start many workflows in parallel, running each workflow
over a longer period of time on only a few VMs to minimize cost. In comparison,
the dynamic algorithms tend to run one workflow at a time across many VMs in
parallel.

@@

To evaluate and compare the three proposed algorithms, we developed a cloud
workflow simulator~\cite{CloudWorkflowSimulator} based on CloudSim~\cite{Calheiros2011}. Our
simulation model consists of Cloud, VM and WorkflowEngine entities. The Cloud entity starts and
terminates VM entities using an Amazon EC2-like API. VM entities simulate the
execution of individual tasks, including randomized variations in runtime. The
WorkflowEngine entity manages the scheduling of tasks and the provisioning of 
VMs based on the chosen algorithm. We assume that the VMs have a single core
and execute tasks sequentially. 
Although CloudSim provides a more advanced infrastructure model, which includes
time-sharing and space-sharing policies, we do not use these features since we
are interested mainly in the execution of tasks on VMs and high-level workflow
scheduling and provisioning. 
The simulator reads workflow description files in
a modified version of the DAX format used by the Pegasus Workflow Management
System~\cite{Deelman2005}. 
%The modified file format was created for the
%synthetic workflow generator developed by Bharathi, et al~\cite{Bharathi2008}.

@@

Pareto ensembles contain a small number of larger workflows and a large number
of smaller workflows. Their sizes  
%in a Pareto ensemble 
are chosen according to a Pareto distribution. The distribution was modified so
that the number of large workflows (of size $\geq$ 900) is increased by a small amount to
produce a ``heavy-tail''. This causes Pareto ensembles to have a slightly larger
number of large workflows, which reflects behavior commonly observed in many
computational workloads. 
An example of the distribution of workflow sizes that
occurs in a Pareto ensemble is shown in Figure~\ref{fig:ensemble-pareto}.

\begin{figure}[t] 
    \centering
    \includegraphics[width=0.6\columnwidth]{figures/ensemble-pareto}
    \caption[Histogram of workflow sizes in Pareto ensembles]{Histogram of
    workflow sizes in Pareto ensembles. Workflow size is measured in number of
    tasks.}
    \label{fig:ensemble-pareto}
\end{figure}

@@

\begin{figure*}[tb]
    \centering
    \includegraphics[width=\textwidth]{run-finish-variations-test-0-output-distributions}
    \caption{Percentage of high scores achieved by each algorithm on different ensemble types for all five applications. C = Constant ensembles, PS = Pareto Sorted ensembles, PU = Pareto Unsorted ensembles, US = Uniform Sorted ensembles, and UU = Uniform Unsorted ensembles.} 
    \label{fig:distributions}
\end{figure*}

@@

Figure~\ref{fig:distributions} shows the percentage of simulations for which
each algorithm achieved the highest score for a given ensemble type. This
experiment was conducted using all five applications, with all five types of
ensembles. For each application and ensemble type, 10 random ensembles of 50
workflows each were created. Each ensemble was simulated with all three
algorithms using 10 budgets and 10 deadlines (1,000 simulations per
application, ensemble type, and algorithm, or 75,000 in total). The best scores percentage is
computed by counting the number of times that a given algorithm achieved the
highest score and dividing by 1000. Note that it is possible for multiple
algorithms to get the same high score (to tie), so the numbers do not
necessarily add up to 100\%. The sum is much higher than 100\% in cases where
the dynamic algorithms perform relatively well because DPDS and WA-DPDS, which
are very similar algorithms, often get the same high score.

There are several interesting things to notice about
Figure~\ref{fig:distributions}. The first is that, in most cases, SPSS
significantly outperforms both dynamic algorithms (DPDS and WA-DPDS). This is
attributed to the fact that SPSS is able to make more intelligent scheduling
and provisioning decisions because it has the opportunity to compare different
options and choose the one that results in the best outcome. In comparison,
the dynamic algorithms are online algorithms and are not able to project into
the future to weigh the outcomes of their choices.

@@

\subsection{Results vs. varying deadline and budget}

\begin{figure*}[htb]
    \centering
    \includegraphics[width=0.9\textwidth]{run-finish-variations-test-0-output-deadlines-app-dist-selected}    
    \caption{Number of completed workflows from the ensemble depending on normalized deadline.}
    \label{fig:completed-deadlines}
\end{figure*}
\begin{figure*}[htb]
    \centering
    \includegraphics[width=0.9\textwidth]{run-finish-variations-test-0-output-budgets-app-dist-selected}    
    \caption{Number of completed workflows from the ensemble depending on normalized budget.}
    \label{fig:completed-budgets}
\end{figure*}

In order to give more details about the performance of the algorithms, in this
section we show how the results depend on the deadline and budget constraints.
As the performance metric we use the number of completed workflows of the
ensemble, since the exponential score is not suitable for showing trends.
We use all deadline and budget parameters as defined in
Section~\ref{sec:exp-parameters} and normalize them to the range from 0 to 1 to
facilitate plotting.
For each normalized deadline, we compute the mean number of completed workflows,
averaged over all budgets and random seeds, so each point on the plot is based
on 100 different simulation runs. In Figure~\ref{fig:completed-deadlines}
and~\ref{fig:completed-budgets} we present the results of CyberShake and LIGO
applications, as representatives of fine-grained and coarse-grained workflows.
Similar results were obtained for other applications.

Figure~\ref{fig:completed-deadlines} shows the results depending on normalized
deadline. It can be observed, that with increasing deadline, the number of
completed worklfows grows steeply and then stabilizes. This means that for
longer deadlines it is usually not possible to complete more workflows due to
the budget constraint. For short deadlines, the algorithms need to allocate more
VM instances in parallel, making it more difficult to utilize the resources
efficiently, which results in lower numbers of completed workflows.

We can also observe that the algorithms are able to complete more workflows in
the ensembles that have priorities assigned randomly (unsorted), in comparison
to the ensembes where the priorities are assigned based on workflow size
(sorted). This results from the fact that in the unsorted distributions the large
workflows often have low priorities, so they can be rejected by the algoritms,
thus making more space in the schedule for smaller high-priority workflows.

One interesting observation from Figure~\ref{fig:completed-deadlines} is that for
CyberShake application the result curves have a sawtooth shape. This results
from the combination of two effects. The first comes from the fact that dynamic
algorithms perform better when the deadline is close to the full hour, due to
the provisioning mechanism that tries to shut down unused VM instances before a
full hour. The second one is that when the deadline increases, it may become
possible to admit a larger workflow with higher priority instead of several
smaller ones with lower priority. That is why the number of completed workflows
may decrese, but the exponential score (utility function) based on priorities
will be higher. These effects are visible for CyberShake application, for which
the maximum deadline is 3 hours, but they become less important for LIGO
application, for which the maximum deadline is 20 hours. See also the discussion
on task granularity in Setion~\ref{sec:granularity}.

@@

@@

\begin{figure*}[htb]
    \centering
%    \subfloat[CyberShake]{
    \includegraphics[width=0.6\textwidth]{run-stretching-test-output-scaling-CYBERSHAKE}
%    }\\
%    \subfloat[Montage]{
    \includegraphics[width=0.6\textwidth]{run-stretching-test-output-scaling-MONTAGE}
%    }
    \caption{Percentage of high scores achieved by each algorithm for CyberShake (top)
    and Montage (bottom) ensembles when task runtime is scaled by a constant factor. A
    scaling factor of $x$ means that the runtime of tasks in each workflow was
    multiplied by $x$.}
    \label{fig:stretching}
\end{figure*}

@@

Figure~\ref{fig:stretching} shows the relative performance of the algorithms as
the scaling factor is increased from 1 to 16 for CyberShake  and Montage
applications. Each data point in the figure represents 500 simulations (5
ensembles with 10 budgets and 10 deadlines). The best scores percent was
calculated the same way as it was for Figure~\ref{fig:distributions}.
